Determinism, Complexity, and Predictability in Computer Performance

### 2014-02-27 results.tex
finished the results section, mostly
moved a small chunk to concl
defined a newcommand
added usepackage{color}

\item that the way information is processed internally by a system is
  correlated with the effectiveness of a given predictor on
  time-series data from that system

@@

The experiments below involve the three different prediction methods
described in Section~\ref{sec:compModel}---na\"ive, ARIMA, and
LMA---applied to time-series data from eight different systems: {\tt
  col\_major}, {\tt 403.gcc}, and the six different regimes of the
{\tt dgesdd} signal in Figure~\ref{fig:wwpe}.  The objective of these
experiments was to explore how prediction accuracy is related to WPE,
and how that relationship depends on the generating process and the
prediction method.  Working from the first 90\% of each signal, we
generated a prediction of the last 10\% of that signal using each of
the methods, as described in Section~\ref{sec:accuracy}, then
calculated the MASE value of those predictions.  We also calculated
the WPE of each time series, as described in
Section~\ref{sec:meaComplex}, using a wordlength of six.  In order to
assess the run-to-run variability of these results, we repeated all of
these calculations on 15 separate trials: i.e., 15 different runs of
each program.  Figure~\ref{fig:wpe_vs_mase_best} shows the {\sl best}
of these 45 predictions for each system: i.e., the lowest error over
all 15 trials and all three methods for each program.  The WPE is
plotted against the corresponding MASE value in order to bring out the
correlation between these two quantities.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.6\textwidth]{figs/prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each system.
% 
% (across all trials and all prediction methods) 
% 
The dashed line is a least-squares linear fit of all the points except
{\tt dgesdd$_1$}, which we have excluded for reasons explained in the
text.  [[Ryan, is that correct?]]}
  \label{fig:wpe_vs_mase_best}
\end{figure}

The best-case prediction error for each of these eight systems is
roughly proportional to the weighted permutation entropy, which is
consistent with our first conjecture.  The dashed line in the figure,
a linear least-squares fit to these best-case errors---with the
exception of {\tt dgesdd$_5$}, for reasons described below---captures
this rough proportionality.  This is not a formal result.  The three
methods used here were chosen to span the space of standard prediction
strategies, but they do not, of course, cover that space exhaustively.
Our goal here is an empirical assessment of predictability and
complexity, not formal results about a ``best'' predictor for a given
time series.  There may be other methods that produce lower MASE
values than those in Figure~\ref{fig:wpe_vs_mase_best}, but the
sparseness of the points in the upper-left and lower-right quadrants
of this plot strongly suggests that the underlying predictability of a
time series is inversely correlated with its WPE.  The rest of this
section describes our results in more detail, including the measures
taken to assure meaningful comparisons across methods, trials, and
programs, and elaborates on the meaning of the dashed line in the
figure.

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for all 360
experiments.
\begin{figure}
  \centering
    \includegraphics[width=0.6\textwidth]{figs/ARIMA_prediction_vs_entropy}
    \includegraphics[width=0.6\textwidth]{figs/ARIMA_prediction_vs_entropy}
    \includegraphics[width=0.6\textwidth]{figs/LMA_prediction_vs_entropy}
\caption{WPE vs. MASE for all trials, methods, and systems.  \svdone,
  \svdthree, and \svdfive are omitted from the top plot for scale
  reasons; their MASE scores are $2.6763\pm4.3282$, $31.3857\pm
  0.2820$, and $20.8703 \pm 0.1915$ respectively.  {\color{red}Replace
    the top plot with a plot of the naive results---but running from 0
    to 3.1 on the horizontal axis and excluding SVD 1, 3, and 5}.  The
  dashed line is the same as in the previous figure [[Ryan, is this
      correct?]].  Numerical values, including means and standard
  deviations of the errors, can be found in Table~\ref{tab:error} in
  the appendix.
% 
%  [[All traces except those from \svd$_1$ lie above the line,
%      indicating that LMA is better suited prediction method for the
%      computer traces considered.]]  [[This comes {\bf long} before the
%      associated discussion in the text, and it's not what this figure
%      is trying to show here.]]
% 
}
    \label{fig:wpe_vs_mase_all}
\end{figure} 
There are 15 points in each cluster, one for each trial\footnote{A
  trial involves collecting a time series from the corresponding
  system, predicting the last 10\% of that time series, and
  calculating the MASE score.}.  (The points in
Figure~\ref{fig:wpe_vs_mase_best} are the leftmost points in any of
the three clusters for the corresponding system.)  Most of the
clusters are quite tight.  Some are flat and wide because the WPE
values of each run of any particular program tend to have little
variance; for most traces, the LMA and ARIMA variance is low as well,
resulting in small, tight clusters.  The {\sl spatial distribution} of
the clusters is quite different in the three plots: LMA is tight
around the dashed line, whereas naive and ARIMA are [[need to fill in
    words here when get the third image]].  This is a geometric
validation of the second conjecture in this paper, as described in the
following paragraphs.

Though \col is a very simple program, its dynamics are actually quite
complicated, as discussed in Section~\ref{sec:intro}.  Recall from the
right-hand column of Figure~\ref{fig:gcc_vs_col} that the naive and
ARIMA prediction methods perform poorly on this signal (MASE values
$0.57 \pm$ 0.001 and $0.59 \pm 0.21$, respectively).  That is, the
errors produced by these methods are a little over half as large as
the errors produced by the random-walk method, which simply uses the
current value as the prediction.  However, the WPE value for the \col
trials is $0.51 \pm 0.003$: i.e., right in the middle of the
complexity spectrum described in Section~\ref{sec:intro}, implying
that roughly half of the signal is predictive structure.  This
disparity---WPE values that suggest a high rate of forward information
transfer in the signal, but predictions with low MASE
scores---manifests geometrically in the top two images in
Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far to
the right of the dashed line.  This is important; it says that the
methods aren't leveraging that information very well.  This is not
surprising, really.  The dynamics of \col may be complicated, but they
are not unstructured.  This signal is nonlinear, deterministic, and
chaotic.  Indeed, if you use a prediction technique that uses a
nonlinear model (LMA) you get a MASE of $0.05 \pm 0.001$---i.e., 20
times more accurate than a random-walk forecast.  Note the position of
the \col cluster in the bottom image in
Figure~\ref{fig:wpe_vs_mase_all}: much closer to the dashed line.
From a practitioner's standpoint, what this means is that being below
that line tells you that there really IS usable structure in the time
series, that you could do better, and that you should use a better
method.  Note, too, that the \col points are clustered in LMA but
spread out horizontally in ARIMA, far to the right of the dashed line
(i.e., much worse than the other methods).  We need to come up with a
hypothesis about this spread and skew, even if it's totally tentative.

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
from this system still contain a significant amount of structure that
can, in theory, be leveraged to predict the future course of the time
series.  However, the effectiveness of any prediction strategy will
depend on how well it leverages the available information.
% 
% (This is the basis for the second conjecture above.)  
% 
This issue becomes apparent in the case of \svdfive, as is clear from
the positions of the corresponding clusters in
Figure~\ref{fig:wpe_vs_mase_all}.  The naive and ARIMA prediction
strategies in the top two plots produce MASE scores of $2.3700 \pm
0.0505 $ and $20.8703 \pm 0.1915$, respectively: that is, 2.37 and
20.87 times worse than a simple random walk forecast of the same
signals.  [[This is big because the amount of variance in this signal
    makes it a bad idea to just guess the mean.  The same thing is
    true of \svdthree.]] [[Joshua: can you say that better---in a way
    that is true of both of these, and distinguishes them from all of
    the signals that have lower values?]] Again, the positions of the
ARIMA and LMA clusters---significantly below and to the right of the
dashed line---should suggest to a practitioner that there is more
structure in this signal than the method is able to leverage, so s/he
can do better and should try a different prediction method.  In this
case, the LMA method produces a MASE score of $ 0.7177\pm 0.0483 $ and
a cluster of results on the WPE-MASE plot that is much closer to the
dashed line.  This validates our second conjecture: the naive and
ARIMA prediction methods, which use constant and linear models,
respectively, cannot capture or reproduce the way in which the
\svdfive system processes information, but the nonlinear LMA method
can.

The WPE of \gcc is much higher: $0.94 \pm 0.001$.  This system
transmits very little information forward in time and provides almost
no structure for prediction methods to work with.  What we see: naive
does the best, then LMA, then ARIMA.  Hypothesis: that that 5\% of the
structure is nonlinear, so LMA can use it better than ARIMA can.
Naive does better than LMA because it's filtering out the noise.

\svdone behaves very differently than the other seven systems in this
study: though its WPE is very high ($0.95 \pm 0.02$), two of the three
prediction methods (LMA and ARIMA) do quite well: $0.82 \pm 0.08$ and
$0.7141 \pm 0.07$ MASE scores, respectively.  Accordingly, the
clusters of \svdone points in the bottom two images in
Figure~\ref{fig:wpe_vs_mase_all} are well above the trend followed by
the other seven clusters.  Also, the MASE scores of the predictions
produced by the na\"ive method are highly inconsistent: $2.67 \pm
4.33$.  These effects, we believe, are artifacts of the way MASE is
calculated.  Recall that MASE scores are scaled relative to a random
walk forecast, which uses the current value as the prediction.  This
strategy works very badly on signals with frequent, large, rapid
transitions.  Consider a signal that oscillates from one end of its
range to the other at every step.  A signal like this will have a low
WPE, much like \col.  However, a random-walk forecast of this signal
will be 180 degrees out of phase with the true continuation.  Since
random-walk error appears in the denominator of the MASE score, this
effect can shift points leftwards on a WPE vs. MASE plot, and that is
exactly why the \svdone clusters for ARIMA and LMA are above the
dashed line in Figure~\ref{fig:wpe_vs_mase_all}.  The \svdone time
series, which is shown in Figure~\ref{fig:svdone-ts},
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\textwidth]{figs/svdonets2}
\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
\end{figure} 
is not quite to the level of the worst-case signal described above,
but it poses a serious challenge to random-walk prediction.  It is
dominated by a white-noise regime (between 1.86 and 1.88 on the
vertical scale in the Figure), punctuated by short excursions above
1.9.  In the former regime, which makes up more than 80\% of the
signal, there are frequent dips to 1.82 and occasional larger dips
below 1.8.  These single-point dips are the bane of random walk
forecasting.  In this particular case, roughly 40\% of the forecasted
points are off by the width of the associated dip, which skews the
associated MASE scores.  Signals like this are also problematic for
the naive prediction strategy, since the outliers have significant
pull on the mean, which compounds the effect of the skew in the
scaling factor and creates a large spread in the \svdone MASE values.
For all of these reasons, we view \svdone as an outlier and thus
exclude it from the fit calculation of the dashed line in
Figure~\ref{fig:wpe_vs_mase_best}.

@@

Comparing Figures~\ref{fig:lma_pred_vs_ent} and
\ref{fig:arima_pred_vs_ent} also elucidates another key finding:
usable and quantifiable predictive structure can be present in a time
series without a prediction scheme being able to utilize it. In
particular, information may be transferred from past to future through
the present but because of the mechanism the underlying process uses
to process that information (e.g., linear or nonlinear) certain
prediction strategies may be blind to or not be able to efficiently
utilize this information. For example, consider \col, programs like
this have been shown to exhibit deterministic chaos
\cite{mytkowicz09}. If this were the case with \col, an out-of-the-box
linear method like ARIMA would simply be ill-equipped to model and
utilize the kind of structure present, as is evident in Figure
\ref{fig:arima_pred_vs_ent}. In contrast, a nonlinear predictor like
LMA which is built to handle deterministic chaos, can interpret and
utilize this type of structure just fine. We believe that many of the
shifts in forecast accuracy for low-to-moderate WPE programs between
ARIMA and LMA is precisely happening for this reason: Just because
there is forward information transfer, does not mean that an arbitrary
predictor can interpret or utilize it, but luckily WPE can tell us
when this structure is present as shown in Figure
\ref{fig:lma_vs_arima}.

@@

The big picture: Furthermore, since LMA can predict nonlinear behavior
while ARIMA cannot, we see that the clusters in Figure
~\ref{fig:arima_pred_vs_ent} are mostly further to the right than
those in Figure~\ref{fig:lma_pred_vs_ent}.

>>>>

\item that the way information is generated and processed by a system
  is correlated with the effectiveness of a given predictor on
  time-series data from that system

@@

The experiments below involve three different prediction methods
applied to time-series data from eight different systems: {\tt
  col\_major}, {\tt 403.gcc}, and the six different regimes of the
{\tt dgesdd} signal in Figure~\ref{fig:wwpe}.  The objective of these
experiments was to explore how prediction accuracy is related to WPE,
and how that relationship depends on the generating process and the
prediction method.  Working from the first 90\% of each signal, we
generated a prediction of the last 10\% using the \naive, ARIMA, and
LMA prediction methods, as described in Section~\ref{sec:accuracy},
then calculated the MASE value of those predictions.  We also
calculated the WPE of each time series using a wordlength of six, as
described in Section~\ref{sec:meaComplex}.  In order to assess the
run-to-run variability of these results, we repeated all of these
calculations on 15 separate trials: i.e., 15 different runs of each
program.  Figure~\ref{fig:wpe_vs_mase_best} shows the {\sl best} of
these 45 predictions for each system: i.e., the lowest error over all
15 trials and all three methods for each of the eight programs.  The
WPE is plotted against the corresponding MASE value here in order to
bring out the correlation between these two quantities.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.6\textwidth]{figs/prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each system.
% 
% (across all trials and all prediction methods) 
% 
The dashed line is a least-squares linear fit of all the points except
{\tt dgesdd$_1$}, which we have excluded for reasons explained in the
text.  [[Ryan, is that correct?]]}
  \label{fig:wpe_vs_mase_best}
\end{figure}

The best-case prediction error for each of these eight systems is
roughly proportional to the weighted permutation entropy, which is
consistent with our first conjecture.  The dashed line in the figure,
a linear least-squares fit to these best-case errors\footnote{with the
  exception of {\tt dgesdd$_1$}, for reasons described below} captures
this rough proportionality.  This is not a formal result.  The three
methods used here were chosen to span the space of standard prediction
strategies, but they do not cover that space exhaustively.  Our goal
here is an empirical assessment of predictability and complexity, not
formal results about a ``best'' predictor for a given time series.
There may be other methods that produce lower MASE values than those
in Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
in the upper-left and lower-right quadrants of this plot strongly
suggests that the underlying predictability of a time series is
inversely correlated with its WPE.  The rest of this section describes
our results in more detail, including the measures taken to assure
meaningful comparisons across methods, trials, and programs, and
elaborates on the meaning of the dashed line in the figure.

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for all 360
experiments.
\begin{figure}
  \centering
    \includegraphics[width=0.6\textwidth]{figs/ARIMA_prediction_vs_entropy}
    \includegraphics[width=0.6\textwidth]{figs/ARIMA_prediction_vs_entropy}
    \includegraphics[width=0.6\textwidth]{figs/LMA_prediction_vs_entropy}
\caption{WPE vs. MASE for all trials, methods, and systems.  \svdone,
  \svdthree, and \svdfive are omitted from the top plot for scale
  reasons; their MASE scores are $2.676 \pm 4.328$, $31.386 \pm
  0.282$, and $20.870 \pm 0.192$ respectively.  {\color{red}Replace
    the top plot with a plot of the naive results---but running from 0
    to 3.1 on the horizontal axis and excluding SVD 1, 3, and 5}.  The
  dashed line is the same as in the previous figure [[Ryan, is this
      correct?]].  Numerical values, including means and standard
  deviations of the errors, can be found in Table~\ref{tab:error} in
  the appendix.
% 
%  [[All traces except those from \svd$_1$ lie above the line,
%      indicating that LMA is better suited prediction method for the
%      computer traces considered.]]  [[This comes {\bf long} before the
%      associated discussion in the text, and it's not what this figure
%      is trying to show here.]]
% 
}
    \label{fig:wpe_vs_mase_all}
\end{figure} 
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding system in any of these three plots.)  The WPE
values do not vary very much across trials.  For most traces, the
variance in MASE scores is low as well, resulting in small, tight
clusters.  In some cases---ARIMA predictions of {\tt col\_major}, for
instance---the MASE variance is larger, which spreads out the clusters
horizontally.  The distribution of the cluster {\sl positions} is
quite different in the three plots: clusters of predictions generated
with the nonlinear LMA method are closer to the dashed line, whereas
the naive and ARIMA prediction clusters are more spread out [[need to
    revisit this wording when get the third image]].  This is a
geometric validation of the second conjecture in this paper, as
described in the following paragraphs.

Though \col is a very simple program, its dynamics are actually quite
complicated, as discussed in Section~\ref{sec:intro}.  Recall from the
graphical representations in the right-hand column of
Figure~\ref{fig:gcc_vs_col} that the \naive and ARIMA prediction
methods do not perform as well on this signal as the nonlinear LMA
method.  The MASE values of the \naive and ARIMA predictions are $0.57
\pm$ 0.001 and $0.59 \pm 0.21$, respectively, across all 15 trials.
That is, the errors produced by these methods are a little over half
as large as the errors produced by the random-walk method---a
primitive strategy that simply uses the current value as the
prediction.  However, the WPE value for the \col trials is $0.51 \pm
0.003$, which is in the center of the complexity spectrum described in
Section~\ref{sec:intro}.  [[This suggests that roughly half of the
    signal is predictive structure.]]  [[I'm still uncomfortable with
    that phrasing; let's revisit it after we have the terminology and
    definitions hammered out.]]

This disparity---WPE values that suggest a high rate of forward
information transfer in the signal, but predictions with comparatively
poor MASE scores---is obvious in the geometry of the top two images in
Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far to
the right of the dashed line.  {\sl This indicates that these methods
  are not leveraging the available information in the signal.}  The
dynamics of \col may be complicated, but they are not unstructured.
This signal is nonlinear and deterministic, and if one uses a
prediction technique that is based a nonlinear model (LMA)---rather
than a method that simply predicts the running mean (\naive) or one
that uses a linear model (ARIMA)---the MASE score is much better:
$0.050 \pm 0.001$.  This prediction is 20 times more accurate than a
random-walk forecast, which is more in line with the level of
predictive structure that the low WPE value suggests is present in the
signal.  The dashed line is a heuristic that captures this argument.
It can allow practitioners to recognize when a prediction method is
not well matched to the task at hand: that is, when the time series
has more predictive structure than the method is able to use.  When
the predictions are well below that line (viz., the top images in
Figure~\ref{fig:wpe_vs_mase_all}), one should try a different method.

Note that the \col points are clustered in LMA but spread out
horizontally in ARIMA, far to the right of the dashed line (i.e., much
worse than the other methods).  [[We need to come up with a hypothesis
    about this spread and skew, even if it's totally tentative.  We
    may want to merge this discussion into one of the other \col
    paragraphs---or perhaps put it with the \svdone discussion later
    on?]]

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
from this system still contain a significant amount of structure that
can, in theory, be leveraged to predict the future course of the time
series.  As mentioned above, though, in the discussion of {\tt
  col\_major}, the effectiveness of any prediction strategy will
depend on how well it leverages the available information; methods
that use linear models, for instance, may fail when applied to
nonlinear processes.
% 
% (This is the basis for the second conjecture above.)  
% 
The match between model and task is an issue in the \svdfive
experiments as well.  The MASE scores of the \naive and ARIMA
predictions for this system were $20.870 \pm 0.192$ and $2.370 \pm
0.051$, respectively: that is, 20.87 and 2.37 times worse than a
simple random walk forecast of the same signals\footnote{The \naive
  MASE score is large because the amount of variance in this signal is
  high, which makes guessing the mean a particularly bad strategy.
  The same thing is true of the \svdthree signal.}.  Again, the
positions of these points on a WPE vs. MASE plot---significantly below
and to the right of the dashed line---should suggest to a practitioner
that there is more structure in this signal than the corresponding
method is able to leverage, and that a different method might do
better.  Indeed, for \svdfive, the LMA method produces a MASE score of
$ 0.718\pm 0.048 $ and a cluster of results that is much closer to the
dashed line on the WPE-MASE plot.  Again, this validates our second
conjecture: the LMA method can capture and reproduce the way in which
the \svdfive system processes information, but the \naive and ARIMA
prediction methods cannot.

The WPE of \gcc is much higher: $0.940 \pm 0.001$.  This system
transmits very little information forward in time and provides almost
no structure for prediction methods to work with.  [[What we see:
    \naive does the best, then LMA, then ARIMA.  Hypothesis: that that
    5\% of the structure is nonlinear, so LMA can use it better than
    ARIMA can.  \naive does better than LMA because it's filtering out
    the noise.]]

\svdone behaves very differently than the other seven systems in this
study: though its WPE is very high ($0.95 \pm 0.02$), two of the three
prediction methods do quite well, yielding MASE scores of $0.82 \pm
0.08$ (LMA) and $0.714 \pm 0.07$ (ARIMA).  This pushes the
corresponding clusters of points in the bottom two images in
Figure~\ref{fig:wpe_vs_mase_all} well above the trend followed by the
other seven clusters.  Also, the MASE scores of the predictions of
this system produced by the \naive method are highly inconsistent:
$2.67 \pm 4.33$.  These effects, we believe, are artifacts of the way
MASE is calculated.  Recall that MASE scores are scaled relative to a
random-walk forecast, which uses the current value as the prediction.
This strategy works very badly on signals with frequent, large, rapid
transitions.  Consider a signal that oscillates from one end of its
range to the other at every step.  A signal like this will have a low
WPE, much like {\tt col\_major}.  However, a random-walk forecast of
this signal will be 180 degrees out of phase with the true
continuation.  Since random-walk error appears in the denominator of
the MASE score, this effect can shift points leftwards on a WPE
vs. MASE plot, and that is exactly why the \svdone clusters for ARIMA
and LMA are above the dashed line in Figure~\ref{fig:wpe_vs_mase_all}.
This time series, which is shown in closeup in
Figure~\ref{fig:svdone-ts},
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\textwidth]{figs/svdonets2}
\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
\end{figure} 
is not quite to the level of the worst-case signal described above,
but it still poses a serious challenge to random-walk prediction.  It
is dominated by a white-noise regime (between $\approx$1.86 and
$\approx$1.88 on the vertical scale in the Figure), punctuated by
short excursions above 1.9.  In the former regime, which makes up more
than 80\% of the signal, there are frequent dips to 1.82 and
occasional larger dips below 1.8.  These single-point dips are the
bane of random-walk forecasting.  In this particular case, roughly
40\% of the forecasted points are off by the width of the associated
dip, which skews the associated MASE scores.  Signals like this are
also problematic for the \naive prediction strategy, since the
outliers have significant influence on the mean.  This compounds the
effect of the skew in the scaling factor and creates a large spread in
the \svdone MASE values.  For all of these reasons, we view \svdone as
an outlier and exclude it from the fit calculation of the dashed line
in Figure~\ref{fig:wpe_vs_mase_best}.

@@

@@
